---
title: "Principles of good graphics"
author: "Therese Anders"
date: "5/6/2019"
output:
  slidy_presentation:
    font_adjustment: +1
    footer: "Therese Anders (tanders@usc.edu) - FSU Summer Methods School 2019"
    incremental: true
    highlight: pygments
    fig_width: 7
    fig_height: 6
    fig_caption: true
---


# Day 1: Basics of data visualization

1. Principles of good graphics
2. The grammar of graphics
3. Building a graph in `ggplot2`
4. Primer on data management using `dplyr` and `tidyr`

# Day 2: Descriptive data visualization

1. Changing the appearance of charts
  * Labels
  * Themes
  * Colors
  * Scales
2. Scatterplots
3. Heatmaps

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

